Windowed least-squares model reduction for dynamical systems

### 2020-10-09 main.tex
latest updates

The \methodAcronym\ approach can be viewed as a hybrid space--time method and displays 
commonalities with several related efforts. 
First, Ref.~\cite{bui_thesis} briefly formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically it mentioned the
optimize-then-discretize vs.\ discretize-then-optimize approaches. This work
did not fully develop this approach, eschewing it 
for 
sequential residual minimization in time (i.e., LSPG). The present work thus
formally develops and extends several of the concepts put forth
in Ref.~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1) this work also considers trial subspaces characterized by spatial dimension reduction only, and (2) we
minimize the time-continuous FOM residual over arbitrary time
windows. We note that,
similar to the current work, Ref.~\cite{constantine_strom} employs
minimization of the time-continuous FOM residual as a starting point; as such,
this work shares some thematic similarities with
Ref.~\cite{constantine_strom}.  Lastly, Ref.~\cite{choi_stlspg} develops a
space--time extension of LSPG projection that minimizes the time-discrete
FOM residual over the entire time domain.  The present work distinguishes
itself from Ref.~\cite{choi_stlspg} in that (1) this work minimizes the
time-continuous FOM residual, (2) this work minimizes this residual over
arbitrary time windows, and (3) this work also considers trial subspaces
associated with spatial dimension reduction only.

@@

\subsection{Reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. Two types of space--time trial subspaces are commonly used for
	this purpose:\footnote{For both spatial and space--time ROMs of dynamical systems, all trial subspaces are, strictly speaking, space--time subspaces.} 
\begin{enumerate} 
	\item \textit{Subspaces that reduce only the spatial dimension of the full-order
		model (\spatialAcronym)}. These trial subspaces are characterized by a spatial projection operator, associate with a basis that represents the spatial dependence of the state, and are employed in classic model reduction approaches, e.g., Galerkin and LSPG. %These 
%	trial subspaces are commonly used
%		in model-reduction methods based on spatial projection followed by time
%		integration (e.g., the Galerkin and LSPG approaches).
	\item \textit{Subspaces that reduce both the spatial and temporal dimensions of the full-order
		model (\spaceTimeAcronym)}.
These trial subspaces are characterized by a space--time projection operator, associate with a basis that represents the spatial and temporal dependence of the state, and are employed in space--time 
model reduction approaches (e.g., space--time Galerkin~\cite{benner_st}, space--time LSPG~\cite{choi_stlspg}). 
% These trial subspaces are commonly employed in
%		model-reduction methods based on space--time projection (e.g., space--time
%		LSPG).
\end{enumerate}
 We now describe these two types of space--time trial subspaces and their
	application to the Galerkin, LSPG, and space--time LSPG approaches.

@@

\begin{appendices}
\section{Proper orthogonal decomposition}
Algorithm~\ref{alg:pod} presents the algorithm for computing the trial basis via proper orthogonal decomposition.
\begin{algorithm}
\caption{Algorithm for generating POD Basis.}
\label{alg:pod}
\textbf{Input:} Number of time-steps between snapshots $N_{\text{skip}}$; intercept $\stateInterceptArg{}$; basis dimension $K$ \; 
%\newline
\textbf{Output:} POD Basis $\basisspace \in \mathbb{V}_{K}(\RR{N})$ \;
%\newline
\textbf{Steps:}
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect solutions into snapshot matrix
$$\mathbf{S}(N_{\text{skip}}) \defeq \begin{bmatrix} \stateFOMDiscreteArg{0} - \stateInterceptArg{n} & \stateFOMDiscreteArg{N_{\text{skip}}} - \stateInterceptArg{n} & \cdots & \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} - \stateInterceptArg{n} \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$\mathbf{S} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
    \item Truncate left singular vectors and to form the basis, $\basisspace \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_K \end{bmatrix}$
\end{enumerate}

@@

@@

\begin{algorithm}
\caption{Algorithm for generating the sampling matrix through q-sampling.}
\label{alg:qdeim}
\textbf{Input:} Number of time-steps between snapshots $N_{\text{skip}}$, number of primal sampling points, $n_s$ \; 
%\newline
\textbf{Output:} Weighting matrix $\stweightingMatArg{} \equiv \stweightingMatOneTArg{} \stweightingMatOneArg{} \in \{0,1\}^{N \times N}$ \;
%\newline
\textbf{Steps:}
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect velocity snapshots 
$$\mathbf{F}(N_{\text{skip}}) \defeq \begin{bmatrix} \velocity (\stateFOMDiscreteArg{0})  & \velocity( \stateFOMDiscreteArg{N_{\text{skip}}}) & \cdots & \velocity( \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} ) \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$\mathbf{F} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
    \item Compute the QR factorization of  $\mathbf{U}^T$ with column pivoting,
    \begin{equation*}
        \mathbf{U}^T \mathbf{P}^* = \mathbf{QR}
    \end{equation*}
    with $\mathbf{P}^* \equiv \begin{bmatrix} \mathbf{p}_1 & \cdots & \mathbf{p}_{\text{floor}(N_t/N_{\text{skip}})} \end{bmatrix}$, $\mathbf{p}_i \in \{0,1\}^N$. 
    \item Select the first $n_s$ columns of $\mathbf{P}^*$ to form the sampling point matrix $\stweightingMatOneTArg{} \in \{0,1\}^{N \times n_s}$. 
    \item Augment the sampling point matrix, $\stweightingMatOneArg{}$, with additional columns such that all unknowns are computed at the mesh cells selected by Step 3. For the second numerical experiment, these additional unknowns correspond to each conserved variable and quadrature point in the selected cells. 
\end{enumerate}

>>>>

The \methodAcronym\ approach can be viewed as a hybrid space--time method and displays 
commonalities with several related efforts. 
First, Ref.~\cite{bui_thesis} briefly formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically it mentioned the
optimize-then-discretize vs.\ discretize-then-optimize approaches. This work
did not fully develop this approach, eschewing it 
for 
sequential residual minimization in time (i.e., LSPG). The present work thus
formally develops and extends several of the concepts put forth
in Ref.~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1) this work also considers trial subspaces characterized by spatial dimension reduction only, and (2) we
minimize the time-continuous FOM residual over arbitrary time
windows. We note that,
similar to the current work, Ref.~\cite{constantine_strom} employs
minimization of the time-continuous FOM residual as a starting point; as such,
this work shares some thematic similarities with
Ref.~\cite{constantine_strom}.  Additionally, Ref.~\cite{choi_stlspg} develops a
space--time extension of LSPG projection that minimizes the time-discrete
FOM residual over the entire time domain.  The present work distinguishes
itself from Ref.~\cite{choi_stlspg} in that (1) this work minimizes the
time-continuous FOM residual, (2) this work minimizes this residual over
arbitrary time windows, and (3) this work also considers trial subspaces
associated with spatial dimension reduction only. Lastly, we note that the windowing approach proposed here  
displays commonalities with space--time methods employing space--time slabs, e.g., Refs.\cite{HUGHES1988339,PIRONNEAU1992117,HUGHES1996217,HUGHES1989173}. 
In these approaches, the space--time domain is typically decomposed into a series of space--time slabs. A system of algebraic equations is then formed over each slab 
via, e.g., the discontinuous Galerkin method, and the space--time method operates by sequentially marching through these space--time slabs. From this perspective, the proposed windowing approach is best viewed as a space--time method in where each window comprises multiple time-slabs. The time-slabs within each window experience a two-way coupling to each other, by virtue of the residual minimization statement being defined over the entire window, and thus one can no longer sequentially march over the time-slabs.

@@

\subsection{Reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. Two types of space--time trial subspaces are commonly used for
	this purpose:\footnote{For both spatial and space--time ROMs of dynamical systems, all trial subspaces are, strictly speaking, space--time subspaces.} 
\begin{enumerate} 
	\item \textit{Subspaces that reduce only the spatial dimension of the full-order
		model (\spatialAcronym)}\footnote{In this work, we refer to the ``state-space" dimension of an ordinary differential equation as the ``spatial" dimension. We use this notation due to the common application of projection-based ROMs to semi-discretized partial differential equations. We emphasize, however, that the presented approach is applicable to ordinary differential equations that do not associate with some type of spatial discretization, i.e., as in molecular dynamics.}
. These trial subspaces are characterized by a spatial projection operator, associate with a basis that represents the spatial dependence of the state, and are employed in classic model reduction approaches, e.g., Galerkin and LSPG. %These 
%	trial subspaces are commonly used
%		in model-reduction methods based on spatial projection followed by time
%		integration (e.g., the Galerkin and LSPG approaches).
	\item \textit{Subspaces that reduce both the spatial and temporal dimensions of the full-order
		model (\spaceTimeAcronym)}.
These trial subspaces are characterized by a space--time projection operator, associate with a basis that represents the spatial and temporal dependence of the state, and are employed in space--time 
model reduction approaches (e.g., space--time Galerkin~\cite{benner_st}, space--time LSPG~\cite{choi_stlspg}). 
% These trial subspaces are commonly employed in
%		model-reduction methods based on space--time projection (e.g., space--time
%		LSPG).
\end{enumerate}
 We now describe these two types of space--time trial subspaces and their
	application to the Galerkin, LSPG, and space--time LSPG approaches.

@@

\begin{appendices}
\section{Proper orthogonal decomposition}
Algorithm~\ref{alg:pod} presents the algorithm for computing the trial basis via proper orthogonal decomposition.
\begin{algorithm}
\caption{Algorithm for generating POD basis.}
\label{alg:pod}
\textbf{Input:} Number of time-steps between snapshots $N_{\text{skip}}$; intercept $\stateInterceptArg{}$; basis dimension $K$; weighting matrix, $\mathbf{P}$ \; 
%\newline
\textbf{Output:} POD Basis $\basisspace \in \{ \mathbf{X} \in \RR{\fomdim \times \romdim} | \mathbf{X}^T\mathbf{P} \mathbf{X} = \mathbf{I} \} $ \;
%\newline
\textbf{Steps:}
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect solutions into snapshot matrix
$$\mathbf{S}(N_{\text{skip}}) \defeq \begin{bmatrix} \stateFOMDiscreteArg{0} - \stateInterceptArg{n} & \stateFOMDiscreteArg{N_{\text{skip}}} - \stateInterceptArg{n} & \cdots & \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} - \stateInterceptArg{n} \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$ \sqrt{\mathbf{P}}^T\mathbf{S} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$ and $\sqrt{\mathbf{P}}$ is the Cholesky decomposition of $\mathbf{P}$, i.e., $\sqrt{\mathbf{P}} \sqrt{\mathbf{P}}^T = \mathbf{P}$
    \item Truncate left singular vectors and to form the basis, $\basisspace \equiv \begin{bmatrix} [\sqrt{\mathbf{P}}^T]^{-1}  \mathbf{u}_1 & \cdots & [\sqrt{\mathbf{P}}^T]^{-1} \mathbf{u}_K \end{bmatrix}$
\end{enumerate}

@@

%In the final numerical experiment, time-local trial bases are constructed for a set of varying input parameters. Algorithm~\ref{alg:pod_tlp} describes the construction of the POD bases for this case.
%\begin{algorithm}
%\caption{Algorithm for generating time-local POD basis over multiple parameter instances}
%\label{alg:pod_tlp}
%\textbf{Input:} Number of time-steps between snapshots $N_{\text{skip}}$; starting time-instance, $N_s$; number of instances $N_{i}$; intercept $\stateInterceptArg{}$; parameter instances $\param_{1} , \ldots, \param_{n_{\text{train}}}$; basis dimension $K$ \\
%%\newline
%\textbf{Output:} POD Basis $\basisspace \in \mathbb{V}_{K}(\RR{N})$ \;
%%\newline
%\textbf{Steps:}
%\begin{enumerate}
%    \item Solve FOM O$\Delta$E for $\param_1, \ldots, \param_{n_{\text{train}}}$ and collect solutions into snapshot matrix
%\small{$$\mathbf{S}(N_{\text{skip}}) \defeq \begin{bmatrix} \stateFOMDiscreteParamArg{N_s}{\param_1} - \stateInterceptArg{n} & \cdots & \stateFOMDiscreteParamArg{N_s + N_i N_{\text{skip}}}{\param_1}- \stateInterceptArg{n} & \cdots & \stateFOMDiscreteParamArg{N_s + N_i N_{\text{skip}}}{\param_{n_{\text{train}}}}- \stateInterceptArg{n} \end{bmatrix}$$}
%    \item Compute the thin singular value decomposition, $$\mathbf{S} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
%    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
%    \item Truncate left singular vectors and to form the basis, $\basisspace \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_K \end{bmatrix}$
%\end{enumerate}
%\end{algorithm}

@@

\begin{algorithm}
\caption{Algorithm for generating the sampling matrix through q-sampling.}
\label{alg:qdeim}
\textbf{Input:} Number of time-steps between snapshots $N_{\text{skip}}$, number of primal sampling points, $n_s$ \; 
%\newline
\textbf{Output:} Weighting matrix $\stweightingMatArg{} \equiv \stweightingMatOneTArg{} \stweightingMatOneArg{} \in \{0,1\}^{N \times N}$ \;
%\newline
\textbf{Steps:}
\begin{enumerate}
    \item Solve FOM O$\Delta$E and collect velocity snapshots 
%$$\mathbf{F}(N_{\text{skip}}) \defeq \begin{bmatrix} \velocity (\stateFOMDiscreteArg{0})  & \velocity( \stateFOMDiscreteArg{N_{\text{skip}}}) & \cdots & \velocity( \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} ) \end{bmatrix}$$
$$\mathbf{F}(N_{\text{skip}}) \defeq \begin{bmatrix} \stateFOMDiscreteArg{0} & \stateFOMDiscreteArg{N_{\text{skip}}}  & \cdots & \stateFOMDiscreteArg{\text{floor}(N_t/N_{\text{skip}})N_{\text{skip}}} \end{bmatrix}$$
    \item Compute the thin singular value decomposition, $$\mathbf{F} (N_{\text{skip}}) = \mathbf{U \Sigma Z}^T,$$
    where $\mathbf{U} \equiv \begin{bmatrix} \mathbf{u}_1 & \cdots & \mathbf{u}_{\text{floor}(N_t/N_{\text{skip}})}\end{bmatrix}$
    \item Compute the QR factorization of  $\mathbf{U}^T$ with column pivoting,
    \begin{equation*}
        \mathbf{U}^T \mathbf{P}^* = \mathbf{QR}
    \end{equation*}
    with $\mathbf{P}^* \equiv \begin{bmatrix} \mathbf{p}_1 & \cdots & \mathbf{p}_{\text{floor}(N_t/N_{\text{skip}})} \end{bmatrix}$, $\mathbf{p}_i \in \{0,1\}^N$. 
    \item Select the first $n_s$ columns of $\mathbf{P}^*$ to form the sampling point matrix $\stweightingMatOneTArg{} \in \{0,1\}^{N \times n_s}$. 
    \item Augment the sampling point matrix, $\stweightingMatOneArg{}$, with additional columns such that all unknowns are computed at the mesh cells selected by Step 3. For the second numerical experiment, these additional unknowns correspond to each conserved variable and degree of freedom in the selected cells. 
\end{enumerate}
